Supplemental Materials for ``Universals of word order reflect optimization of grammars for efficient communication''

### 2019-10-11 writeup/pnas-word-order-si.tex
code

For each language, we extracted all lemmas of words heading an \emph{xcomp} dependency, occurring at least 2 times.
In 39 languages, all extracted words had the same dominant order.
Additionally, in four Germanic languages (Afrikaans, Danish, Dutch, and German), the verb of volition (Afrikaans \emph{wil}, Danish \emph{ville}, Dutch \emph{willen}, German \emph{wollen}) is mostly annotated with the \emph{aux} relation due to UD annotation guidelines, but in all languages, its dominant order (verb of volition before its complement) agrees with the dominant order of the \emph{xcomp} dependency (head-initial).
In three historical languages (Ancient Greek, Latin, and Old Church Slavonic), verbs of volition agree with the dominant order of \emph{xcomp}, while several other verbs that do not indicate volition show opposite dominant order.
Finally, in Gothic, the verb of volition (\emph{wiljan}) has opposite dominant order, resulting in an apparent violation of Correlation \raisebox{.5pt}{\textcircled{\raisebox{-.9pt} {8}}}.

@@

%TODO
%$\sum_{t \in C} \log p(t|u) \approx |C| H[P|T]  $
%$\sum_{t \in C} \sum_{i} \log p(t|u) \approx |C| H[P|T]  $

@@

TODO also make a plot with the 0.0 ad 0.9 lines

\begin{figure}
\centering
\includegraphics[width=\textwidth]{../results/plane/analyze_pareto_optimality/pareto-plane-perLanguage-arrows-smoothed-halfspace-untransformed.pdf}
	\caption[]{TODO}\label{fig:lambda-halfplane-09}
\end{figure}

@@

@@

For computational reasons, we restrict the vocabulary to the most frequent 50,000 words in the treebanks for a given language.
Given the moderate size of the corpora, this limit is only attained only for few languages.
In each time step, the input is a concatenation of embeddings for the word, for language-specific POS tags, and for universal POS tags.
The model predicts both the next word and its POS tags in each step.
Using POS tags is intended to prevent overfitting on small corpora.

TODO explain.
Here we replicate Study 1.
We

\begin{figure}
\centering
\includegraphics[width=\textwidth]{../results/plane/unlexicalized/pareto-plane-perLanguage-lexicalized-mle.pdf}
	\caption[Predictability and Parseability]{Study 1, replication with lexicalized models: Predictability and parseability of 51 languages, for \emph{lexicalized} models, compare Figure~\ref{fig:pareto-per-lang}.}
\end{figure}

\begin{table}
\centering
\small{
\begin{tabular}{l||ll|lll|lll}
Language & Pred. (t) & Parse. (t) & \multicolumn{3}{c|}{Pred. (Binomial)} & \multicolumn{3}{c}{Parseab. (Binomial)} \\ 
&  $p$ & $p$ &  Est. &CI & $p$ & Est. & CI & $p$  \\ \hline \hline
\input{../results/plane/unlexicalized/per-language-statistics-unlexicalized.tex}
\end{tabular}
}
	\caption{Study 1, replication with lexicalized models: Per-language results in Study 1, with \emph{lexicalized} parsers and word-level-only language models. Compare Table~\ref{fig:pareto-per-lang-stats}}
\end{table}

\begin{figure}
	\centering
	\includegraphics[width=\textwidth]{../results/plane/unlexicalized/analyze_pareto_optimality/figures/quantileByLambda-mle.pdf}
	\caption{Study 1, replication with lexicalized models: Optimality by $\lambda$ (lexicalized models).}
\end{figure}

>>>>

For each language, we extracted all lemmas of words heading an \emph{xcomp} dependency, occurring at least 10 times.
In 39 languages, all extracted words had the same dominant order.
Additionally, in four Germanic languages (Afrikaans, Danish, Dutch, and German), the verb of volition (Afrikaans \emph{wil}, Danish \emph{ville}, Dutch \emph{willen}, German \emph{wollen}) is mostly annotated with the \emph{aux} relation due to UD annotation guidelines, but in all languages, its dominant order (verb of volition before its complement) agrees with the dominant order of the \emph{xcomp} dependency (head-initial).
In three historical languages (Ancient Greek, Latin, and Old Church Slavonic), verbs of volition agree with the dominant order of \emph{xcomp}, while several other verbs that do not indicate volition show opposite dominant order.
Finally, in Gothic, the verb of volition (\emph{wiljan}) has opposite dominant order, resulting in an apparent violation of Correlation \raisebox{.5pt}{\textcircled{\raisebox{-.9pt} {8}}}.

@@

%$\sum_{t \in C} \log p(t|u) \approx |C| H[P|T]  $
%$\sum_{t \in C} \sum_{i} \log p(t|u) \approx |C| H[P|T]  $

@@

\begin{figure}
\centering
\includegraphics[width=\textwidth]{../results/plane/analyze_pareto_optimality/pareto-plane-perLanguage-arrows-smoothed-halfspace-untransformed.pdf}
	\caption[]{Per-Language results as in Figure~\ref{fig:pareto-per-lang-raw}, with lines connecting those points that have the same efficiency as the real grammar at $\lambda = 0.0$ (dotted) and $\lambda = 0.9$ (dashed). Points to the bottom/left of these line have lower efficiency than the real grammar, at the given value of $\lambda$.}\label{fig:lambda-halfplane-09}
\end{figure}

@@

For computational reasons, we restrict the vocabulary to the most frequent 50,000 words in the treebanks for a given language.
Given the moderate size of the corpora, this limit is only attained only for few languages.
In each time step, the input is a concatenation of embeddings for the word, for language-specific POS tags, and for universal POS tags.
The model predicts both the next word and its language-specific POS tag in each step.
Using POS tags is intended to prevent overfitting on small corpora.
This choice was made before evaluating the efficiency of real grammars, and before evaluating word order properties.

Here we replicate Study 1 with both parsers and language models operating entirely on word forms, without POS tags.
Results are shown in Figures~\ref{fig:pareto-plane-noPOS}.
As shown in Figure~\ref{fig:byLambda-noPOS}, most real grammars have higher efficiency than most baselines across permissible values of $\lambda$.

\begin{figure}
\centering
\includegraphics[width=\textwidth]{../results/plane/unlexicalized/pareto-plane-perLanguage-lexicalized-mle.pdf}
	\caption[Predictability and Parseability]{Study 1, replication with lexicalized models: Predictability and parseability of 51 languages, for \emph{lexicalized} models, compare Figure~\ref{fig:pareto-per-lang}.}\label{fig:pareto-plane-noPOS}
\end{figure}

\begin{table}
\centering
\small{
\begin{tabular}{l||ll|lll|lll}
Language & Pred. (t) & Parse. (t) & \multicolumn{3}{c|}{Pred. (Binomial)} & \multicolumn{3}{c}{Parseab. (Binomial)} \\ 
&  $p$ & $p$ &  Est. &CI & $p$ & Est. & CI & $p$  \\ \hline \hline
\input{../results/plane/unlexicalized/per-language-statistics-unlexicalized.tex}
\end{tabular}
}
	\caption{Study 1, replication with lexicalized models: Per-language results in Study 1, with \emph{lexicalized} parsers and word-level-only language models. Compare Table~\ref{fig:pareto-per-lang-stats}}\label{tab:pareto-plane-noPOS}
\end{table}

\begin{figure}
	\centering
	\includegraphics[width=\textwidth]{../results/plane/unlexicalized/analyze_pareto_optimality/figures/quantileByLambda-mle.pdf}
	\caption{Study 1, replication with lexicalized models: Optimality by $\lambda$ (lexicalized models).}\label{fig:byLambda-noPOS}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=\textwidth]{../results/plane/pureUD/pareto-plane-perLanguage-pureUD-mle.pdf}
	\caption[Predictability and Parseability]{Study 1, replication with the original UD format models: Predictability and parseability of 51 languages, for \emph{lexicalized} models, compare Figure~\ref{fig:pareto-per-lang-pureUD}.}
\end{figure}

\begin{figure}
	\centering
	\includegraphics[width=\textwidth]{../results/plane/pureUD/analyze_pareto_optimality/figures/quantileByLambda-mle.pdf}
	\caption{Study 1, replication with original UD format: Optimality by $\lambda$ (lexicalized models).}
\end{figure}

In Study 1, we considered deterministic ordering grammars, even though real languages have some degree of word order freedom.
Here, we show that deterministic baseline grammars provide conservative baselines: They have higher efficiency than baseline grammars with word order freedom comparable to real languages.

In order to obtain baselines whose freedom is comparable to that of real languages, we constructed baselines that have the same Branching Direction Entropy \cite{futrell2015quantifying}:
For every one of the 37 syntactic relations, we chose $a_\tau$ the direction entropy to be the same as observed in the actual orderings found in the UD corpus.
We kept the relative ordering of siblings on one side deterministic, i.e., we considered the limit where the values $b_\tau$ for different $\tau$ are far apart.

Results are shown in Figure~\ref{fig:nondeterministic-english}.
For every one of the baseline grammars, we show both its deterministic and its nondeterministic version.
Nondeterministic grammars are strongly less efficient than deterministic grammars, showing that deterministic grammars constitute stronger baselines.

\begin{figure}
	\centering
	\includegraphics[width=0.4\textwidth]{../results/plane/nondeterministic/pareto-plane-perLanguage-nondeterministic-mle.pdf}
	\caption{Parseability and predictability, including nondeterministic versions of the 50 baseline grammars.}\label{fig:nondeterministic-english}
\end{figure}

@@
